---
title: "Analyzing Socioeconomic and Housing Factors to Predict Arrest Trends in New York State"
author: "Joyce Yan (qy249), Lydia Lin (dl2253), Victoria Xiao (sx287)"
date: today
format: pdf
---

# Set up

```{r}
#| label: import-packages

library(dplyr)
library(readr)
library(ggplot2)
library(reshape2)
library(tidyverse)
library(tidymodels)
library(poissonreg)
```

# Objective

Our project aims to predict **the total number of adult arrests in New
York State counties based on key social, economic, and housing factors
from 2015 to 2022**. By analyzing historical data and identifying
patterns, we want to understand how these factors influence arrest
counts and use this knowledge to create a reliable predictive model.

This model can help policymakers and law enforcement anticipate trends
in crime and allocate resources more effectively. Our goal is to address
two key problems:

1.  Identifying which socioeconomic and housing factors are most
    strongly associated with arrest counts.
2.  Developing a predictive tool to estimate future arrest counts, which
    can guide decisions on crime prevention and community support.

By achieving these objectives, we hope to contribute to data-driven
strategies for building safer, more equitable communities.

# Data description

Our analysis-ready dataset combines arrest data with socioeconomic and
housing characteristics for counties in New York State from 2015 to
2022. The dataset is organized at the county-year level and includes
both arrest counts by offense type and demographic, economic, and
housing indicators as rates. This structure makes it suitable for
exploring relationships and building predictive models.

## Variables in the dataset

### General Information

-   **County**: The name of the county in New York State.
-   **Year**: The year of the data, ranging from 2015 to 2022.

### Arrest Data

-   **Total**: Total adult arrest counts in the county for the
    respective year.
-   **Felony Total**: Total felony arrests.
    -   **Drug Felony**: Arrests for drug-related felonies.
    -   **Violent Felony**: Arrests for violent felonies.
    -   **DWI Felony**: Arrests for driving while intoxicated (DWI)
        felonies.
    -   **Other Felony**: Arrests for other felony offenses.
-   **Misdemeanor Total**: Total misdemeanor arrests.
    -   **Drug Misdemeanor**: Arrests for drug-related misdemeanors.
    -   **DWI Misdemeanor**: Arrests for DWI misdemeanors.
    -   **Property Misdemeanor**: Arrests for property-related
        misdemeanors.
    -   **Other Misdemeanor**: Arrests for other misdemeanor offenses.

### Social Characteristics

-   **Less_than_9th_grade**: Percentage of the population aged 25 and
    over with less than a 9th-grade education.
-   **With_disability**: Percentage of the civilian noninstitutionalized
    population with a disability.
-   **Civilian_veterans**: Percentage of the civilian population aged 18
    years and over who are veterans.
-   **Limited_English**: Percentage of the population aged 5 and over
    who speak a language other than English at home and speak English
    less than "very well."

### Housing Characteristics

-   **Total_housing_units**: Total housing units in terms of occupancy
    as a percentage of the population.
-   **With_mortgage**: Percentage of owner-occupied housing units that
    have a mortgage.
-   **Without_mortgage**: Percentage of owner-occupied housing units
    without a mortgage.

### Economic Characteristics

-   **Unemployment_rate**: The percentage of the civilian labor force
    that is unemployed.
-   **Median_household_income**: Median income of all households in the
    county, adjusted for 2022 inflation.
-   **Below_poverty_level**: Percentage of individuals aged 18 and over
    whose income is below the poverty threshold.
-   **No_health_insurance**: Percentage of the civilian
    noninstitutionalized population without health insurance coverage.

The dataset ensures comparability across counties by using rates instead
of raw counts for demographic, economic, and housing variables. This
approach accounts for differences in county population sizes, making the
data suitable for cross-county analysis.

# EDA

Based on our team’s previous exploration, we observed a significant
variation in the population sizes of counties in New York State. Some
counties have substantially larger populations compared to others. To
minimize bias, it would be more appropriate to use percentages for our
feature columns instead of raw counts. Therefore, we decided to revisit
the data cleaning process and transform the count-based features into
percentage-based features. We will then perform exploratory data
analysis using these newly transformed feature columns.

You can find our detailed data cleaning process in the **explore.qmd**
file. I will export a cleaned **merged_data.csv** file with the
percentage features added to facilitate further analysis.

If we were to display all 62 counties in New York State in our EDA, the
plots would become overly cluttered and difficult to interpret.
Therefore, we decided to focus on the top 10 counties with the highest
arrests to initially explore relationships between variables. Later, we
will apply feature selection methods to gain a deeper understanding of
the key features.

```{r}
#| label: merged-data

df = read_csv("data/merged_data.csv")
head(df)
```

## EDA on arrests

```{r}
#| label: eda-arrests

# Determine top 10 counties with the most arrests
top_counties <- df |> 
  group_by(County) |> 
  summarize(Total_Arrests = sum(Total, na.rm = TRUE)) |> 
  arrange(desc(Total_Arrests)) |> 
  slice(1:10) |> 
  pull(County)

# Filter the dataset to include only the top 10 counties
df_top10 <- df |> 
  filter(County %in% top_counties)

# Horizontal bar plot showing the average number of total arrests for each of the top 10 counties
county_summary_top10 <- df_top10 |>
  group_by(County) |>
  summarize(Average_Total_Arrests = mean(Total, na.rm = TRUE))

ggplot(county_summary_top10, aes(x = reorder(County, Average_Total_Arrests), y = Average_Total_Arrests)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  coord_flip() +
  labs(
    title = "Average Arrests Per County (Top 10 Counties)",
    x = "County",
    y = "Average Total Arrests"
  )

# Line plot showing the total number of arrests over time for the top 10 counties
ggplot(df_top10, aes(x = Year, y = Total, color = County)) +
  geom_line() +
  labs(
    title = "Trends in Total Arrests Over Time (Top 10 Counties)",
    x = "Year",
    y = "Total Arrests"
  )

# Grouped bar plot showing the number of arrests by type for each of the top 10 counties
arrests_stacked_top10 <- melt(
  df_top10,
  id.vars = c("County", "Year"),
  measure.vars = c("Drug Felony", "Violent Felony", "DWI Felony", "Other Felony", 
                   "Drug Misdemeanor", "DWI Misdemeanor", "Property Misdemeanor", "Other Misdemeanor"),
  variable.name = "Arrest_Type",
  value.name = "Count"
)

arrests_felony <- arrests_stacked_top10 |> 
  filter(grepl("Felony", Arrest_Type))

arrests_misdemeanor <- arrests_stacked_top10 |> 
  filter(grepl("Misdemeanor", Arrest_Type))

# Felony
ggplot(arrests_felony, aes(x = Arrest_Type, y = Count, fill = County)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Felony Arrest Counts by Type and County (Top 10 Counties)",
    x = "Felony Arrest Type",
    y = "Total Arrests"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Misdemeanor
ggplot(arrests_misdemeanor, aes(x = Arrest_Type, y = Count, fill = County)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Misdemeanor Arrest Counts by Type and County (Top 10 Counties)",
    x = "Misdemeanor Arrest Type",
    y = "Total Arrests"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Based on our analysis of arrest data in New York State from 2015 to
2022, several key insights emerge. First, urban counties such as Kings,
New York, and Bronx consistently lead in total arrests, with
significantly higher average arrests compared to suburban and rural
counties. This disparity highlights the influence of population density
and socioeconomic challenges on arrest rates. Among felony arrests,
violent crimes contribute the largest proportion after "Other Felonies,"
while property-related offenses account for the majority of misdemeanor
arrests after "Other Misdemeanors." These patterns suggest strong
correlations with socioeconomic factors such as income, unemployment,
and housing instability. Additionally, all counties exhibit a similar
trend in total arrests over time, indicating that year is a significant
factor influencing arrest rates.

## EDA on social characteristics

```{r}
#| label: eda-social

# Reshape social and crime data for scatter plots
social_crime_data <- df_top10 |>
  select(County, Year, Total, Less_than_9th_grade, With_disability, Civilian_veterans, Limited_English) |>
  pivot_longer(cols = c(Less_than_9th_grade, With_disability, Civilian_veterans, Limited_English), 
               names_to = "Social_Variable", values_to = "Social_Value")

# Scatter plot of social variables vs. total crimes
ggplot(social_crime_data, aes(x = Social_Value, y = Total, color = County)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ Social_Variable, scales = "free") +
  labs(
    title = "Relationship Between Social Characteristics and Total Arrests",
    x = "Social Characteristic Value",
    y = "Total Arrests"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Correlation heatmap for social and crime variables
correlation_matrix_social <- df_top10 |>
  select(Less_than_9th_grade, With_disability, Civilian_veterans, Limited_English, 
         Total, `Felony Total`, `Misdemeanor Total`) |>
  cor()

ggplot(melt(correlation_matrix_social), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(
    title = "Correlation Between Social Characteristics and Crime Data",
    x = "",
    y = ""
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), 
    axis.text.y = element_text()
  )
```

The scatter plots reveal a potential positive correlation between social
variables and total arrests, particularly in counties with higher arrest
rates. For instance, counties such as New York and Bronx exhibit both
elevated arrest totals and significant socioeconomic challenges,
including lower education levels and higher rates of disabilities. The
correlation heatmap supports these findings, showing moderate to strong
relationships between social characteristics and crime variables, such
as misdemeanor and felony arrests. These insights suggest that
socioeconomic factors—including education, disability status, and
language barriers—play a critical role in shaping arrest rates.

## EDA on economic characteristics

```{r}
#| label: edd-economic

# Reshape economic and crime data for scatter plots
economic_crime_data <- df_top10 |>
  select(County, Year, Total, Unemployment_rate, Median_household_income, Below_poverty_level, No_health_insurance) |>
  pivot_longer(cols = c(Unemployment_rate, Median_household_income, Below_poverty_level, No_health_insurance), 
               names_to = "Economic_Variable", values_to = "Economic_Value")

# Scatter plot of economic variables vs. total crimes
ggplot(economic_crime_data, aes(x = Economic_Value, y = Total, color = County)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ Economic_Variable, scales = "free") +
  labs(
    title = "Relationship Between Economic Factors and Total Arrests",
    x = "Economic Factor Value",
    y = "Total Arrests"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Correlation heatmap for economic and crime variables
correlation_matrix_economic <- df_top10 |>
  select(Unemployment_rate, Median_household_income, Below_poverty_level, No_health_insurance, 
         Total, `Felony Total`, `Misdemeanor Total`) |>
  cor()

ggplot(melt(correlation_matrix_economic), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(
    title = "Correlation Between Economic Factors and Crime Data",
    x = "",
    y = ""
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), 
    axis.text.y = element_text()
  )
```

The scatter plots highlight notable trends: higher unemployment rates
and greater percentages of the population below the poverty level are
associated with increased total arrests, particularly in counties like
Bronx and New York. Conversely, higher median household incomes show a
negative correlation with total arrests, indicating that wealthier areas
experience fewer arrests. Additionally, the percentage of individuals
without health insurance exhibits a positive correlation with arrest
totals, further emphasizing the role of economic vulnerability in
influencing crime rates. The correlation heatmap reinforces these
findings, with strong positive correlations observed between
unemployment, poverty, and crime variables such as misdemeanor and
felony arrests. These results suggest that economic instability,
including unemployment, poverty, and lack of access to healthcare, are
significant drivers of crime.

## EDA on housing characteristics

```{r}
#| label: eda-housing

# Reshape housing and crime data for scatter plots
housing_crime_data <- df_top10 |>
  select(County, Year, Total, Total_housing_units, With_mortgage, Without_mortgage) |>
  pivot_longer(cols = c(Total_housing_units, With_mortgage, Without_mortgage), 
               names_to = "Housing_Variable", values_to = "Housing_Value")

# Scatter plot of housing variables vs. total crimes
ggplot(housing_crime_data, aes(x = Housing_Value, y = Total, color = County)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ Housing_Variable, scales = "free") +
  labs(
    title = "Relationship Between Housing Characteristics and Total Arrests",
    x = "Housing Characteristic Value",
    y = "Total Arrests"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Correlation heatmap for housing and crime variables
correlation_matrix_housing <- df_top10 |>
  select(Total_housing_units, With_mortgage, Without_mortgage, 
         Total, `Felony Total`, `Misdemeanor Total`) |>
  cor()

ggplot(melt(correlation_matrix_housing), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(
    title = "Correlation Between Housing Characteristics and Crime Data",
    x = "",
    y = ""
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.text.y = element_text() 
  )
```

The scatter plots show no strong trends between housing variables and
total arrests across counties. Similarly, the correlation heatmap
indicates that while total housing units and homes with mortgages have
slight positive correlations with crime metrics, these relationships are
far weaker than those observed with economic or social factors. The
"total housing units" variable forms a straight line, making it
difficult to determine any meaningful trend. Both "with mortgage" and
"without mortgage" variables show slight trends, but it is challenging
to determine whether they are positive or negative, as the scatter
points are widely dispersed. Overall, housing characteristics appear to
have a limited influence on crime patterns, suggesting that other
factors play a more significant role in shaping arrest rates.

# Decisions based on EDA

According to our EDA process, we determined that four economic
variables—"Below Poverty Level," "Median Household Income,"
"Unemployment Rate," and "No Health Insurance"—show a noticeable
correlation with arrests, suggesting that these are critical predictors.
In contrast, the social and housing factors exhibit weaker relationships
with total arrests, with no obvious recognizable trends and low
correlation values in the heatmaps. Instead of outright exclusion, we
decided to perform feature engineering to transform these variables into
potentially more meaningful representations.

```{r}
#| label: feature-engineer

# Create the engineered features
df <- df |>
  mutate(
    # Mortgage Affordability Index
    Mortgage_affordability_index = With_mortgage / (Without_mortgage + 1),

    # Log Housing Units
    Log_housing_units = log(Total_housing_units + 1),
    
    # Log Mortgage Ratio
    mortgage_ratio = With_mortgage / Total_housing_units,
    Log_mortgage_ratio = log(mortgage_ratio + 1),

    # Language × Education Interaction
    Language_education_interaction = Limited_English * Less_than_9th_grade,
    
    # Unemployment × Poverty Interaction
    Unemployment_poverty_interaction = Unemployment_rate * Below_poverty_level
  )

# Filter the dataset to include only the top 10 counties
df_top10 <- df |> 
  filter(County %in% top_counties)

# Reshape engineered data for scatter plots
engineered_data <- df_top10 |> 
  select(County, Year, Total,
         Mortgage_affordability_index, Log_housing_units, 
         Log_mortgage_ratio, Language_education_interaction, 
         Unemployment_poverty_interaction) |> 
  pivot_longer(cols = c(Mortgage_affordability_index, Log_housing_units, 
                        Log_mortgage_ratio, Language_education_interaction, 
                        Unemployment_poverty_interaction), 
               names_to = "Engineered_Variable", values_to = "Engineered_Value")

# Scatter plot of engineered variables vs. total crimes
ggplot(engineered_data, aes(x = Engineered_Value, y = Total, color = County)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ Engineered_Variable, scales = "free") +
  labs(
    title = "Relationship Between Engineered Characteristics and Total Arrests",
    x = "Engineered Characteristic Value",
    y = "Total Arrests"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Correlation heatmap for engineered variables and crime data
correlation_matrix_engineered <- df_top10 |> 
  select(
         Mortgage_affordability_index, Log_housing_units, 
         Log_mortgage_ratio, Language_education_interaction,
         Unemployment_poverty_interaction,
         Total, `Felony Total`, `Misdemeanor Total`) |> 
  cor()

ggplot(melt(correlation_matrix_engineered), aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(
    title = "Correlation Between Engineered Characteristics and Crime Data",
    x = "",
    y = ""
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.text.y = element_text() 
  )

```

## New variables after feature engineering

We performed feature engineering on the dataset to create new derived
features based on domain knowledge and mathematical transformations.

-   **Mortgage_affordability_index**: This feature reflects the relative
    affordability of housing by comparing the number of households with
    a mortgage to those without one. Adding +1 ensures no division by
    zero for areas with zero mortgage-free housing.
-   **Log_housing_units**: This is a logarithmic transformation of the
    total housing units to normalize the data distribution and reduce
    the influence of extreme outliers.
-   **Log_mortgage_ratio**: It applies a logarithmic transformation of
    mortgage ratio to reduce skewness.
-   **Language_education_interaction**: This feature models the
    compounded effect of language barriers and low education levels.
    Combining these two social factors allows us to explore whether
    their combined impact has a stronger association with arrests than
    the individual factors alone.
-   **Unemployment × Poverty Interaction**: Combined "Unemployment Rate"
    and "Below Poverty Level" to account for the compounded effect of
    economic stress on crime rates. This interaction models how the
    combination of unemployment and poverty amplifies arrest rates.

## Features included in models

-   **Mortgage_affordability_index**

-   **Log_housing_units**

-   **Log_mortgage_ratio**

-   **Language_education_interaction**

-   **Unemployment_poverty_interaction**

-   **Unemployment_rate**

-   **Median_household_income**

-   **Below_poverty_level**

-   **No_health_insurance**

-   **Felony Total**

    -   **Drug Felony**
    -   **Violent Felony**
    -   **DWI Felony**
    -   **Other Felony**

-   **Misdemeanor Total**

    -   **Drug Misdemeanor**
    -   **DWI Misdemeanor**
    -   **Property Misdemeanor**
    -   **Other Misdemeanor**

-   **County**

-   **Year**

# Resampling strategy

We chose an **80/20 training-test split** stratified by the target variable, 
Total (total arrest counts), to ensure both subsets reflect the distribution 
of arrest counts. Stratification helps maintain representativeness in both the 
training and testing sets, particularly when the target variable has an uneven 
distribution. This split ratio is standard in predictive modeling and provides 
a good balance between having sufficient data for training and retaining a 
significant portion for unbiased testing.

Additionally, we applied **5-fold cross-validation** within the training set to 
further validate our models. This approach ensures that all data points are used 
for both training and validation, reducing the risk of overfitting while maximizing 
data efficiency. Five folds were selected as a balance between computational 
efficiency and robust evaluation, making it particularly appropriate given the 
size of our dataset and the complexity of our models. 

Together, these strategies 
provide a comprehensive evaluation framework, allowing us to assess model 
generalizability before applying it to the test set.

```{r}
#| label: resampling

# Select relevant columns for modeling
model_features <- df |> 
  select(
    # Socio-economic and housing features
    Mortgage_affordability_index,
    Log_housing_units,
    Log_mortgage_ratio,
    Language_education_interaction,
    Unemployment_poverty_interaction,
    Unemployment_rate,
    Median_household_income,
    Below_poverty_level,
    No_health_insurance,
    
    # Arrest data features
    `Felony Total`,
    `Drug Felony`,
    `Violent Felony`,
    `DWI Felony`,
    `Other Felony`,
    `Misdemeanor Total`,
    `Drug Misdemeanor`,
    `DWI Misdemeanor`,
    `Property Misdemeanor`,
    `Other Misdemeanor`,
    County,
    Year,
    
    # Target variable
    Total
  )

df <- df |> mutate(County = as.character(County))

set.seed(123)

data_split <- initial_split(model_features, prop = 0.8, strata = Total)
train_data <- training(data_split)
test_data <- testing(data_split)

cv_folds <- vfold_cv(train_data, v = 5, strata = Total)

all_metrics <- metric_set(rmse, rsq, mae)
```

# Overview of modeling strategies

## Models to be compared

1.  **Null Model:**

-  Serves as a baseline model for comparison with the performance of 
   Poisson regression and Random Forest.

2.  **Poisson Regression:**

-   We will test two variations:
    -   Using `factor (Year)` to capture year-specific fixed effects.
    -   Using `Year` as a continuous variable to capture linear temporal
        trends.
-   Interaction terms (e.g., **Unemployment Rate × Below Poverty
    Level**) will be included to explore combined effects of key
    predictors.
-   The model’s coefficients will provide insights into the strength and
    direction of relationships between predictors and arrest counts.

3.  **Random Forest:**

-   Train Random Forest models using the same set of predictors as
    Poisson regression:
    -   Economic variables: `Unemployment Rate`,
        `Median Household Income`, `Below Poverty Level`,
        `No Health Insurance`.
    -   Engineered features: `Mortgage Affordability Index`,
        `Log Housing Units`, `Log Mortgage Ratio`,
        `Language × Education Interaction`, and
        `Unemployment × Poverty Interaction`.
    -   Metadata: `County` and `Year`.
-   Tune hyperparameters to optimize performance:
    -   Number of trees (`n_estimators`)
    -   Maximum tree depth
    -   Minimum samples per split

## Evaluation Metrics

-   **Prediction Performance:**
    -   Root Mean Squared Error (RMSE): To evaluate the average
        magnitude of prediction errors.
    -   R-squared (RSQ): To measure how well the model explains the
        variance in the data.
    -   Mean Absolute Error (MAE): To assess the average absolute
        difference between predicted and observed counts.
-   **Variable Importance:**
    -   Use Random Forest’s feature importance scores to identify the
        most influential predictors.

## Training models

### Null model

```{r}
#| label: null-model

# Set up the null model
null_model <- null_model(mode = "regression") |>
  set_engine("parsnip")

# Build the recipe
null_recipe <- recipe(Total ~ ., data = train_data) |>
  step_rm(County)

# Build the workflow
null_workflow <- workflow() |>
  add_model(null_model) |>
  add_recipe(null_recipe)

# Fit the null model to the training data
null_fit <- fit(null_workflow, data = train_data)

# Predict on the testing set
test_predictions <- predict(null_fit, new_data = test_data) |>
  bind_cols(test_data)

# Calculate test metrics
test_metrics <- test_predictions |>
  metrics(truth = Total, estimate = .pred)

test_metrics
```

The RMSE and MAE are significantly large, at 17,257.99 and 12,393.07,
reflecting the simplicity of the null model. The R-squared value is NA,
which is expected for a null model, as it does not use any predictors
and cannot explain the variance in the data. This serves as a baseline
model for comparison with the more complex models that we will create
later.

### Poisson Regression

```{r}
#| label: poisson-regression

base_recipe <- recipe(Total ~ ., data = train_data) |>
  step_mutate(
    Year = as.numeric(Year) 
  ) |>
  step_dummy(all_nominal_predictors(), -all_outcomes()) 

recipe_factor_year <- base_recipe |>
  step_mutate(Year = factor(Year))

model_factor_year <- poisson_reg() |>
  set_engine("glm")

workflow_factor_year <- workflow() |>
  add_recipe(recipe_factor_year) |>
  add_model(model_factor_year)

fit_factor_year <- workflow_factor_year |>
  fit(data = train_data)

recipe_continuous_year <- base_recipe

model_continuous_year <- poisson_reg() |>
  set_engine("glm")

workflow_continuous_year <- workflow() |>
  add_recipe(recipe_continuous_year) |>
  add_model(model_continuous_year)

fit_continuous_year <- workflow_continuous_year |>
  fit(data = train_data)

# Evaluate the models on the test set
predictions_factor_year <- fit_factor_year |>
  predict(new_data = test_data) |>
  bind_cols(test_data)

predictions_continuous_year <- fit_continuous_year |>
  predict(new_data = test_data) |>
  bind_cols(test_data)

metrics_factor_year <- predictions_factor_year |>
  metrics(truth = Total, estimate = .pred)

metrics_continuous_year <- predictions_continuous_year |>
  metrics(truth = Total, estimate = .pred)

# Metrics for Poisson Regression with Year factor
print(metrics_factor_year)

# Metrics for Poisson Regression with Year as a continuous variable
print(metrics_continuous_year)
```

The analysis shows that using factor(Year) in the Poisson regression
model provides better predictive accuracy compared to treating Year as a
continuous variable. The model with factor(Year) achieves lower RMSE and
MAE values and a slightly higher R², indicating it captures
year-specific variations and trends more effectively. In contrast, the
model using Year as a continuous variable struggles to account for
non-linear or year-specific effects, resulting in less accurate
predictions. These results suggest that including factor(Year) is a
better approach for modeling total arrests when using Posisson
regression.

### Random Forest

```{r}
#| label: random-forest

# Feature engineering: focusing on transformations
base_recipe <- recipe(Total ~ ., data = train_data) |>
  step_mutate(
    Year = as.numeric(Year), # Convert Year to numeric
    Mortgage_affordability_index = log(Mortgage_affordability_index + 1)
  ) |>
  step_dummy(all_nominal_predictors(), -all_outcomes()) |> # Dummy encode categorical variables
  step_zv(all_predictors()) |> # Remove zero-variance predictors to avoid issues during modeling
  step_normalize(all_numeric_predictors(), -all_outcomes()) # Normalize all numeric features

rf_spec <- rand_forest(
  mode = "regression",
  trees = tune(),
  mtry = tune(),
  min_n = tune()
) |>
  set_engine("ranger", importance = "impurity")

rf_workflow <- workflow() |>
  add_recipe(base_recipe) |>
  add_model(rf_spec)

rf_grid <- grid_regular(
  trees(range = c(500, 1500)),
  mtry(range = c(3, 12)),
  min_n(range = c(3, 10)),
  levels = 5
)

set.seed(123)

rf_res <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(rmse, rsq, mae)
)

best_rf <- select_best(rf_res, metric = "rmse")

final_rf_workflow <- finalize_workflow(rf_workflow, best_rf)

final_rf_fit <- final_rf_workflow |> fit(data = train_data)

rf_predictions <- final_rf_fit |>
  predict(new_data = test_data) |>
  bind_cols(test_data)

rf_metrics <- rf_predictions |>
  metrics(truth = Total, estimate = .pred)

print(rf_metrics)
```

The Random Forest Model has shown improvement, achieving an RMSE of
868.47 and an R² of 0.9975, showing very good model fit and predictive
accuracy. The MAE value of 304.18 further highlights the model's
improved capability to closely predict the actual arrest values. By
exploring the feature importance, we determined that crime-specific
features, such as violent felonies and total felonies, are the strongest
drivers of arrest counts. While less impactful, housing and
mortgage-related variables contribute additional context to the model.
They may indirectly capture socioeconomic conditions that correlate with
crime levels.

# Model evaluation

The objective of this project was to predict the total number of adult arrests 
in New York State counties based on key social, economic, and housing factors 
from 2015 to 2022. By building predictive models, we aimed to understand how these 
factors influence arrest counts, ultimately helping policymakers make informed 
decisions about crime prevention and resource allocation.

We evaluated three models: the **Null Model**, the **Poisson Regression Model**, 
and the **Random Forest Model**.

1.  **Null Model:** The null model served as a baseline for comparison. 
It included no predictors and simply predicted the mean value of the 
target variable (total number of adult arrests).

|            |               |              |
|------------|---------------|--------------|
| **Metric** | **Estimator** | **Estimate** |
| RMSE       | Standard      | 17,257.99    |
| R²         | Standard      | NA           |
| MAE        | Standard      | 12,393.07    |

- **RMSE:** The high RMSE of 17,257.99 indicates poor predictive accuracy, as the 
null model ignores any variability in the data.
- **R²:** Not applicable (NA) because the null model does not include predictors.
- **MAE:** The MAE of 12,393.07 further highlights the model’s inability to account 
for important predictors.

Overall, the null model demonstrates that using just the mean of the target 
variable is insufficient for accurate predictions and underscores the need 
for more complex models.

2.  **Poisson Regression Model:** We applied a Poisson regression model
    with two variations:

- Variation 1 used factor (Year) to capture year-specific fixed effects.
- Variation 2 used Year as a continuous variable to capture linear trends.
We observed that Variation 1 provided better predictive accuracy, so we
report those results here.

|            |               |              |
|------------|---------------|--------------|
| **Metric** | **Estimator** | **Estimate** |
| RMSE       | Standard      | 459.         |
| R²         | Standard      | 0.999        |
| MAE        | Standard      | 224.         |

- **RMSE:** The RMSE of 459 shows a substantial improvement over the null model, 
reflecting much better predictive accuracy.
- **R²:** The model explains 99.9% of the variance in arrest counts, highlighting 
its ability to capture key factors influencing arrest trends.
- **MAE:** The MAE of 224 indicates a significant reduction in average prediction
error compared to the null model.

The strong performance of the Poisson regression model, particularly in accounting for 
year-specific effects, demonstrates the importance of temporal factors in arrest trends.

3.  **Random Forest Model:** The Random Forest model was trained using key predictors 
from the dataset. Hyperparameter tuning was conducted to optimize its performance.

|            |               |              |
|------------|---------------|--------------|
| **Metric** | **Estimator** | **Estimate** |
| RMSE       | Standard      | 868.47       |
| R²         | Standard      | 0.9975       |
| MAE        | Standard      | 304.18       |

- **RMSE:** The RMSE of 868.47 is higher than the Poisson regression 
model’s RMSE of 459 but still represents a significant improvement over 
the null model.
- **R²:** With an R² of 0.9975, the Random Forest model explains 99.75% of the variance in arrest counts, which is competitive with the Poisson regression model.
- **MAE:** The MAE of 304.18 is higher than the Poisson regression model’s MAE of 224, suggesting that the Poisson regression achieves better accuracy for predicting individual arrest counts.

While the Random Forest model exhibits a slightly higher RMSE and MAE compared to the Poisson regression model, its R² value demonstrates that it effectively captures overall variance in arrest counts.

# Insights and Conclusion

The evaluation of three models highlighted key insights into predicting arrest trends 
in New York State counties. The **Null Model**, serving as a baseline, relied solely on the 
mean of the arrest data and exhibited high errors, underscoring its insufficiency for 
accurately modeling arrest trends. In contrast, the **Poisson Regression Model** demonstrated 
the best overall performance, with an RMSE of 459 and an R² of 0.999, capturing 99.9% 
of the variance in arrest counts. By incorporating year-specific fixed effects, it effectively 
captured temporal patterns, emphasizing the importance of accounting for year-to-year 
variability. This model is particularly suited for understanding overarching trends and 
guiding policy decisions on resource allocation.

We selected the **Poisson Regression Model** as the final model for deployment due to its 
strong predictive performance and interpretability. Its ability to reliably capture 
temporal dynamics, coupled with its low RMSE, makes it the most appropriate choice for 
predicting arrest trends over time. While the **Random Forest Model** achieved competitive 
results, particularly with a lower MAE of 304.18 for individual-level predictions, the 
Poisson Regression Model’s ability to explain broader trends and provide actionable 
insights was prioritized.

# Recommendations

Based on the Poisson Regression Model’s results, we recommend leveraging its predictions 
to guide policy decisions aimed at reducing arrest rates. The model’s ability to capture 
temporal patterns highlights the importance of year-to-year trends, suggesting that 
policymakers should monitor changes in socioeconomic and housing factors over time. 
Specifically, resources could be allocated to counties with the highest predicted 
arrest counts, focusing on addressing the underlying social and economic issues 
driving these trends. The model can also serve as a decision-support tool to evaluate 
the potential impact of proposed interventions, such as community programs or housing 
initiatives, by incorporating changes in key predictors and assessing their effects 
on arrest rates.

However, the model has limitations that should be addressed in future improvements. 
For example, the Poisson Regression Model assumes a linear relationship between 
predictors and the logarithm of the outcome, which may oversimplify complex interactions 
between variables. Additionally, incorporating more predictors, such as crime policies 
or law enforcement practices, could enhance the model’s accuracy and provide a more 
comprehensive understanding of arrest patterns. Since the final model does not achieve 
an exceptionally low RMSE, it may be beneficial to create hybrid models by combining 
the Poisson model with Random Forest predictions. This approach could leverage the 
strengths of both models to achieve better accuracy.
